---
title: "USGS Data"
author: "Gavin McNicol"
date: "2/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Exploring available USGS data (mostly water) for candidate McNicol Lab wetland / wet prairie field sites.

First, load packages:
```{r}
library(tidyverse)
library(waterData)
```

Get help page for the `importDVs` function:

```{r}
??importDVs
```

The format is: `importDVs(staid, code = "00060", stat = "00003", sdate = "1851-01-01", edate = as.Date(Sys.Date(), format = "%Y-%m-%d"))`
  
Note that:
    - Code **00060**: stream flow in cubic feet per second
    - Stat **00003**: mean, for mean daily stream flow

## Blackberry Creek USGS Stream Data

1. **Blackberry Creek**
    + Station 1: "Jericho Lake"

```{r}
jericho <- importDVs( "05551675" , code = "00060", stat = "00003", sdate = "1851-01-01",
  edate = as.Date(Sys.Date(), format = "%Y-%m-%d"))
str(jericho)
```

Try plotting all the data:

```{r}
jericho %>% 
  ggplot(aes(dates, val, color = qualcode)) + 
  geom_line() +
  labs(x = "Year", y = "Stream Flow (cubic ft per s)") +
  theme_classic()
```

The record runs from 1998 to present and almost all the data got the best quality flag. Only 10 missing values.

Does stream flow go to zero in winter? Let's look at the last year of winter data (Dec - March 2019-2020).

```{r}
jericho_winter <- importDVs( "05551675" , code = "00060", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(jericho_winter)
```

```{r}
jericho_winter %>% 
  ggplot(aes(dates, val, color = qualcode)) + 
  geom_line() +
  labs(x = "Year", y = "Stream Flow (cubic ft per s)") +
  theme_classic()
```

It appears some amount of flow continues during the winter.

Let's check the **water temperature** values to see if it was every subzero (when you might expect no flow).

```{r}
jericho_winter_temps <- importDVs( "05551675" , code = "00010", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(jericho_winter_temps)
```

Looks like *water temperatures are not available* at Jericho. What about **dissolved O2**?

```{r}
jericho_winter_o2 <- importDVs( "05551675" , code = "00300", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(jericho_winter_o2)
```

Nope, it appears it's only stream flow and gage height (stage, in feet). This was actually mentioned in the USGS station information. 

1. **Blackberry Creek**
    + Station 2: "N Yorkville"

```{r}
yorkville <- importDVs( "05551700" , code = "00060", stat = "00003", sdate = "1851-01-01",
  edate = as.Date(Sys.Date(), format = "%Y-%m-%d"))
str(yorkville)
```

Try plotting:

```{r}
yorkville %>% 
  ggplot(aes(dates, val, color = qualcode)) + 
  geom_line() +
  labs(x = "Year", y = "Stream Flow (cubic ft per s)") +
  theme_classic()
```

This record is much longer. 1960-present. It captures a large flood in the late 1990s. Almost all the data also got the highest quality flag. Only 6 missing values.

Let's ask the same questions at Yorkville.
Does stream flow go to zero in winter? Let's look at the last year of winter data (Dec - March 2019-2020).

```{r}
yorkville_winter <- importDVs( "05551700" , code = "00060", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(yorkville_winter)
```

```{r}
yorkville_winter %>% 
  ggplot(aes(dates, val, color = qualcode)) + 
  geom_line() +
  labs(x = "Year", y = "Stream Flow (cubic ft per s)") +
  theme_classic()
```

It appears some amount of flow continues during the winter.

Let's check the **water temperature** values to see if it was every subzero (when you might expect no flow).

```{r}
yorkville_winter_temps <- importDVs( "05551700" , code = "00010", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(yorkville_winter_temps)
```

Looks like *water temperatures are not available* at yorkville. What about **dissolved O2**?

```{r}
yorkville_winter_o2 <- importDVs( "05551700" , code = "00300", stat = "00003", sdate = "2019-12-01",
  edate = as.Date("2020-03-31", format = "%Y-%m-%d"))
str(yorkville_winter_o2)
```

Nope, it also appears it's only stream flow and gage height (stage, in feet). 

## Blackberry Creek USGS Well Data

There is one head water well with data from 2013-2016. Look at [this planning document](https://docs.google.com/document/d/1wwahbof9F4_AHNwjYMf4ATcnPpM91vaSCr03aHPOYDQ/edit#). It appears to show the water table rising above the surface by up to 1-2 ft in the winter and dropping in the summer. The summer water table for 2012 looks much lower than 2013. Check to see if the discharge data reflect this:

```{r}
yorkville <- importDVs( "05551700" , code = "00060", stat = "00003", sdate = "2012-01-01",
  edate = as.Date("2015-12-31", format = "%Y-%m-%d"))
str(yorkville)
```

Try plotting:

```{r}
yorkville %>% 
  ggplot(aes(dates, val, color = qualcode)) + 
  geom_line() +
  labs(x = "Year", y = "Stream Flow (cubic ft per s)") +
  theme_classic()
````

Yes, it appears 2012 had a much drier winter than 2013, 2014 or 2015. (Though we lack the corresponding water table data points for 2014 and 2015).

